---
title: "Exercise 5"
author: "Hanna Julku"
output: html_document
---

# 5: Dimentionality reduction techniques

## Human data set

```{r}
# load libraries to be used
library(tidyverse)
library(corrplot)
library(GGally)
library(ggplot2)

# read the data into a data frame
human <- read.csv(file = "\\\\ad.helsinki.fi/home/h/hkonstar/Desktop/IDOS/IODS-project/data/human.csv", row.names = 1) %>% as.data.frame()

#visualize data and show summary
ggpairs(human)
summary(human)
str(human)
cor(human) %>% corrplot(, type = "upper")
```

There are 8 variables and 155 observations in the data set. 
Variable distributions vary a lot. Some follow normal distribution, others do not. 
Life.Exp and Edu.Exp as well as Mat.Mor and Ado.Birth have the highest positive correlation. Life.Exp and Mat.Mor on the hand have highest negative correlation. Just mentioned two "clusters" (Life.Exp + Edu.Exp and Mat.Mor + Ado.Birth) correlate negatively with each other. 

## Principal component analysis (PCA)

```{r}
# perform principal component analysis (with the SVD method)
pca_human <- prcomp(human)

# Show the variability captured by the principal components.
summary(pca_human)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human, choices = 1:2, cex = c(0.8, 1), col = c("grey40", "deeppink2"))
```

Data was not standardized, so let's standardize it and redo what was just done.

```{r}
# standardize data
human_std <- scale(human) 

# perform principal component analysis (with the SVD method)
pca_human_std <- prcomp(human_std)

# Show the variability captured by the principal components.
summary(pca_human_std)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human_std, choices = 1:2, cex = c(0.8, 1), col = c("grey40", "deeppink2"))
```

Standardization makes it possible to compare variables with each other. 
Standardized results resemble visualized data, whereas non-standardized results were quite different. For instance, non-standardized biplot shows GNI as the most impactfull variable, where as for the standardized data results resemble correlations for original data. Standardized results give more defined results.

Life expectancy, female/male proportion having secondary education, expected year of education, and Gross national income appear together and they have an opposite effect than maternal mortality and number of children being born. These relate to PC1 that explains the most of the variance in the data that is about 54% in this case. 
Female/male proportion on participating in labour and women representation in parliament have similar effect. These explain about 16% of the variance in the data, and they are mostly related to PC2. 

## Tea data set

```{r}
# load tea data set
tea <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/tea.csv", stringsAsFactors = TRUE) %>% as.data.frame()

#explore the data
str(tea)
dim(tea)
view(tea)

#rselect subsets to be drawn
tea1 <- tea[,1:9] %>% as.data.frame()
tea2 <- tea[,10:18] %>% as.data.frame()
tea3 <- tea[,20:29] %>% as.data.frame()
tea4 <- tea[,30:36] %>% as.data.frame()

#visualize  data
pivot_longer(tea1, cols = everything()) %>% 
  ggplot(aes(value)) + facet_wrap("name", scales = "free") + geom_bar() +  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))

pivot_longer(tea2, cols = everything()) %>% 
  ggplot(aes(value)) + facet_wrap("name", scales = "free") + geom_bar() +  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))

pivot_longer(tea3, cols = everything()) %>% 
  ggplot(aes(value)) + facet_wrap("name", scales = "free") + geom_bar() +  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))

pivot_longer(tea4, cols = everything()) %>% 
  ggplot(aes(value)) + facet_wrap("name", scales = "free") + geom_bar() +  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

Data consists of 300 observations and 36 variables. Questionnaire data about how people drink their tea (18 items), how they perceive the product (12 items), and personal details (4 items).

## Multiple correspondence analysis (MCA)

I chose variables Tea, How, sugar, age_Q, and sex.

```{r}
library(FactoMineR)
library(tidyr)

# column names to keep in the dataset
keep_columns <- c("Tea", "How", "sugar", "age_Q", "sex")

# select the 'keep_columns' to create a new dataset
tea_time <- select(tea, keep_columns)

# multiple correspondence analysis
mca <- MCA(tea_time, graph = FALSE)

# summary of the model
summary(mca)

# visualize MCA
plot(mca, invisible=c("ind"), graph.type = "classic", habillage = "quali")
```

People using milk or other things with their tea, favoring green or black tea and being 35-44 or +60 year old have positive effect on both dimension 1 and 2. 
Choose Earl Grey, not adding anything to tea, and being 15-24 years-old have negative effect on dimensions 1 and 2. 

For dim1 gender does not have an effect, but for dim2 there seems to be an opposite effect for gender. 

Among these selected variables there does not seem to be any clusters that could clearly explain the variability in the data.