---
title: "Exercise 4"
author: "Hanna Julku"
output: html_document
---

# 4: Clustering and Classification

## Data set

I will be using the Boston data set from late 1970s, which is about Housing values in suburbs of Boston. First the data set needs to uploaded, it is included in the MASS library.

```{r}
# access the MASS package
library(MASS)

# load the data
data("Boston")

# explore the dataset
str(Boston)
dim(Boston)
```
Data consists of 506 observations and 14 variables. Variables are mostly numerical with 2 integer variables.

Variables include such as crime rate, proportion of owner-occupied old buildings, location related to river, percentage of the lower status population etc.
 
## Graphical overview on the data

```{r}
#access needed libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(tidyverse)
library(GGally)

# create plot matrixes with ggpairs()
p <- ggpairs(Boston[,1:5], lower = list(combo = wrap("facethist", bins = 20)))
p2 <- ggpairs(Boston[,6:10], lower = list(combo = wrap("facethist", bins = 20)))
# draw the plots
p
p2

summary(Boston)
```
Variable distributions differ greatly. In these plots showing half of the relationships, it seems that most of the variables are correlated with each other. 

## Data set standardization

First, I will standardize the data set and view its summaries.
```{r}
# center and standardize variables
boston_scaled <- scale(Boston)

# summaries of the scaled variables
summary(boston_scaled)
```
After scaling mean for each variable is 0, in addition variances for different variables became more uniform.

Now, categorical crime variable will be created and categorical break points are created based on quantiles from crim variable stating per capita crime rate by town. Data set is divided to train and test sets, so that 80% of randomly selected observations belong to train set. 

```{r}
# class of the boston_scaled object
class(boston_scaled)
# this is matrix array, so needs to be changes to to data frame for further analysis
boston_scaled <- as.data.frame(boston_scaled)

# create a quantile vector of crime and print it
bins <- quantile(boston_scaled$crim)
bins

# create a categorical variable 'crime' by using the bins just created as break points
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, label = c("low", "med_low", "med_high", "high"))

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)

# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)


# Creating random selection of the data, so that 80 % will belong to training set, and the rest 20 % to test set
# number of rows in the Boston data set 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]

```

## Linear discriminant analysis

We'll do linear discriminant analysis to test how closely all other variables are located to crime variable in this data set.
The data set was divided into two sets, a train set to teach the model and to test set to test how the model works.
First, we will create the model and plot the results.

```{r}
# linear discriminant analysis
lda.fit <- lda(crime ~ ., data = train)

# print the lda.fit object
lda.fit

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 2)
```

Observations categorized as high seem to cluster together far apart from other variables, even though there are some med_high observations as part of that cluster. Low, med_low, and med_high clusters are partly overlapping, yet they can still be distinguished from each other.
Accessibility to radial high ways (variable rad) seems to have the highest impact on crime category "high".

Further, train model will be used to predict categories in test data.

```{r}
# save the crime categories from test data
correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)

# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
table(correct = correct_classes, predicted = lda.pred$class)
```
Predictions were most accurate for high category, where everything (23) went to the right category.
For med_high about half of the predictions were correct (17), and most mis-categorized were in med_low.
For med_low about 2/3 of the predictions were correct (14) and most mis-categorized predictions were in med_high.
For low less than half of the predictions were correctly located to low category (12) and almost all other predictions were categorized as med_low.
It seems that quite often low and med_high categories are predicted as med_low.

All other variables together are not super good at predicting the crime variable.
This is constant with the lda result from train data. Those above mentioned categories are overlapping and hence their prediction is difficult.

## K-means clustering 

```{r}
# reload the data
data("Boston")
# standardize data
boston_rescaled <- scale(Boston)
```

Distances between observations are calculated by two different methods: euclidean distance matrix and manhattan distance matrix.

```{r}
# euclidean distance matrix
dist_eu <- dist(boston_rescaled)

# look at the summary of the distances
summary(dist_eu)

# manhattan distance matrix
dist_man <- dist(boston_rescaled, method = "manhattan")

# look at the summary of the distances
summary(dist_man)
```
```{r}
# k-means clustering
km <- kmeans(boston_rescaled, centers = 3)

# plot the Boston dataset with clusters
pairs(boston_rescaled[,1:5], col = km$cluster)
pairs(boston_rescaled[,6:10], col = km$cluster)
```

Here, I will investigate what is the optimal number of clusters.

```{r}
set.seed(123)

# determine the number of clusters
k_max <- 8

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(boston_rescaled, k)$tot.withinss}) 

# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line') 
```

Investigation shows that 2 is the optimal number for clusters, as there is an edgy turn on the graph. Let's use that and visualize the clusters.

```{r}
# k-means clustering
km2 <- kmeans(boston_rescaled, centers = 2)

# plot the Boston dataset with clusters
pairs(boston_rescaled[,1:5], col = km2$cluster)
pairs(boston_rescaled[,6:10], col = km2$cluster)
```

For some of the variables cluster colors do seems to fit the clusters nicely. For others, clusters are intermixed with each other and hence not so good at separating observations. When this version is compared with the three cluster version, this seems more tidy.


## 3D plots for train data

Here, I will plot the train data in 3D. I will plot the data twice, first by using crime categories as coloring and the using k-means clusters for coloring using the same number of clusters to see the effect of coloring strategy on the plot.

```{r}
model_predictors <- dplyr::select(train, -crime)
# check the dimensions
dim(model_predictors)
dim(lda.fit$scaling)
# matrix multiplication
matrix_product <- as.matrix(model_predictors) %*% lda.fit$scaling
matrix_product <- as.data.frame(matrix_product)
# Access the plotly package. Create a 3D plot of the columns of the matrix product using the crime classes in train data.
library(plotly)
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color = classes)
# Draw similar plot, but use k-means for setting colors.
# remove the crime variable from train data to run k-means
# use centers = 4 to compare the effect of k-means vs quantile in setting colors
trainkm <- dplyr::select(train, -crime)
kmtrain <- kmeans(trainkm, centers = 4)
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color = kmtrain$cluster)
```

Observations are located similarly in both plots. There is some difference in the coloring of the clusters. Here, I used same number of clusters for both to compare the effect of different clustering method. To summarize, observations are the same in both plots, yet coloring of them is different as quantile is set by the probability of those observations in belonging to particular fraction. K-mean on the other hand tries to minimize the mean distance between geometric points, that is, it is more sensitive for the composition of the data.