---
title: "Assignment 6"
author: "Hanna Julku"
output: html_document
---

# 6: Analysis of longitudinal data

## Data sets

First load the data sets and make sure they are ready to be used.

```{r}
library(tidyverse)

# Read the BPRS and RATS data
BPRSL <- read.table("\\\\ad.helsinki.fi/home/h/hkonstar/Desktop/IDOS/IODS-project/data/BPRSL.csv", sep =",", header = T)

RATSL <- read.table("\\\\ad.helsinki.fi/home/h/hkonstar/Desktop/IDOS/IODS-project/data/RATSL.csv", sep =",", header = T)

# Look at the (column) names of BPRS
names(BPRSL)
# Look at the structure of BPRS
str(BPRSL)
# Print out summaries of the variables
summary(BPRSL)

#do same for RATS
names(RATSL)
str(RATSL)
summary(RATSL)
```

Need to factorize some of the variables and also change the data.frame to tibble.

```{r}
# need to refactor some variables
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

#change to tibble
library(tibble)
BPRSL <- as.tibble(BPRSL)
RATSL <- as.tibble(RATSL)

#check structure
str(BPRSL)
str(RATSL)
```
Now data seems as it was at the end of data wrangling and we are ready to analyze.

## RATS

Let's plot the data for visual inspection.

```{r}
#Access the package ggplot2
library(ggplot2)

# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))

```

Data has not been standardized, so let's do that and then replot. Note, there is one outlier in Group 2, but let's keep that for now.

```{r}
# Standardise the variable Weight
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdweight = scale(Weight)) %>%
  ungroup()

# Plot again with the standardised weight
library(ggplot2)
ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized weight")

```

Results look relatively same as they did before standardizing, yet the scale is different. Let's draw a summary graph.

```{r}
# Number of subjects (per group):
n <- 16

library(dplyr)
library(tidyr)
# Summary data with mean and standard error of Weight by Group and Time 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight)) %>%
  ungroup()

# Plot the mean profiles
library(ggplot2)
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

As expected by the outlier in Group 2, standard error is quite big indicating a lot variability within the values of the group.
Let's still keep that in and draw boxplots for the mean weights for the treatments.

```{r}

library(dplyr)
library(tidyr)
# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline Time 1, indicating firt day of weighting)
RATSL8S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL8S)

# Draw a boxplot of the mean versus treatment
library(ggplot2)
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), days 8-64")
```

Now, let's exclude the outlier in group 2.

```{r}
# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
RATSL8S1 <- filter(RATSL8S, RATSL8S$mean < 580)
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), days 8-64")
```

Mean weight for group two has now about the same variance as the other groups. Value for Group 2 mean is about 50 grams lighter without the outlier.

**T-test and anova**

There are three groups and t-test can only have two. T-tests are done in three sets. Correction for multiple test should be done, but not executed here.

```{r}
# Perform a two-sample t-test
with(RATSL8S1, t.test(mean[Group == 1], mean[Group == 2]))
with(RATSL8S1, t.test(mean[Group == 1], mean[Group == 3]))
with(RATSL8S1, t.test(mean[Group == 2], mean[Group == 3]))
# all comparisons are statistically significant indicating differences in treatment groups

library(dplyr)
library(tidyr)
# Add the baseline from the original data as a new variable to the summary data
RATSL8S2 <- RATSL8S %>%
  mutate(baseline = RATSL$Time[1])

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSL8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)

# 
t <- lm(mean ~ Group, data = RATSL8S1)
anova(t)

```

Results from t-tests and anova indicate that diet group had significant impact on body weight.

## BPRS

Let's plot the data to visualize the effect of treatment on bprs score.

```{r}
library(dplyr)
library(tidyr)
# Check the dimensions of the data
dim(BPRSL)

# Plot the data
library(ggplot2)
ggplot(BPRSL, aes(x = week, y = bprs, group = treatment)) +
  geom_smooth(aes(linetype = treatment)) + scale_x_continuous(name = "week") + scale_y_continuous(name = "bprs (points)") + theme(legend.position = "top")
```

Both treatments seem to decrease the bprs score, indicating severity of symptoms related with schizophrenia. This suggest both treatments alleviate schitzophrenic symptoms.

```{r}
# create a regression model BPRS_reg
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)

# print out a summary of the model
summary(BPRS_reg)
```

**Random intercept model**

```{r}
# access library lme4
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)

```

**Random Intercept and Random Slope Model**

```{r}
# create a random intercept and random slope model
library(lme4)
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)

```

**Random Intercept and Random Slope Model with interaction**

```{r}
# create a random intercept and random slope model with the interaction
library(lme4)
BPRS_ref2 <- lmer(bprs ~ week + treatment + (week | subject) + week*treatment, data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

# draw the plot with the bprs values
ggplot(BPRSL, aes(x = week, y = bprs, group = treatment)) +
  geom_smooth(aes(linetype = treatment)) +
  scale_x_continuous(name = "week") +
  scale_y_continuous(name = "bprs score") +
  theme(legend.position = "top")

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)

library(dplyr)
library(tidyr)
# Create a new column fitted
BPRSL <- BPRSL %>% mutate(Fitted)

# draw the plot with the Fitted values 
library(ggplot2)
ggplot(BPRSL, aes(x = week, y = Fitted, group = treatment)) +
  geom_smooth(aes(linetype = treatment)) +
  scale_x_continuous(name = "week") +
  scale_y_continuous(name = "bprs score") +
  theme(legend.position = "top")
```