--- 
title: "Exercise 2"
output: html_document
--- 

# 2: Regression and Model Validation

```{r}
date()
```

## Dataset
First, read the data into R and inspect it.

```{r}
# access libraries that will be needed at exercise 2
library(tidyverse)
library(ggplot2)
library(GGally)

# let's have a look at the data
students2014 <- read.csv(file = '\\\\ad.helsinki.fi/home/h/hkonstar/Desktop/IDOS/IODS-project/data/exp2data.csv')
str(students2014)
head(students2014)
# if you prefer you may use the following instead or in addition
# glimpse(students2014)
````

This dataset consists of 7 variables and 166 observations.


Variable discriptions:

gender = F for female, M for male

age = age in years

attitude = points as mean for attitudes toward statistics

deep = points as mean for items related with deep approach to studying

stra = points as mean for items related with strategic approach to studying

surf = points as mean for items related with surface approach to studying

points = exam points


### Data Visualization
Visualize the data so that gender is distinguished. Use ggpairs to create matrix of plots from the variables in the dataset.

```{r}

# create a plot matrix with ggpairs()
p <- ggpairs(students2014, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))

# draw the plot
p

# show summaries for variables in students2014
summary(students2014)

````
Visual inspection of the data revealed the following:

There is about double the amount of females in the data in respect to males. 
Scatterplot for points with attitudes indicates a linear interaction. Also for deep and stra there seems to be linear interaction. 

Gender distribution for variables age, deep, and points seem similar. For attitude male distribution is more towards higher scores as for females it reminds more of normal distribution. For stra the case is the opposite for attitude distribution: females are having higher score. For surf females have sharp peaking normal distribution where males are leaning towards lower scores. 

**Numerical summaries for variables:**

Following listing will have 1st and 3rd quadrants and median values for stated variable.

age: 21.0 - 27.0, mean 25.5

attitude: 2.6 - 3.1, median 3.2

deep: 3.3 - 4.1, median 3.7

stra: 2.6 - 3.6, median 3.2

surf: 2.4 - 3.2, median 2.8

points: 19.0 - 27.8, median 23.0


These state the same information as the graph functions.


#### Statistically significant correlations
Surf correlate negatively with attitude, deep, and stra. This means that when surf increases the other one decreases.
When  gender is taken into account stra is not correlated for males nor females. Attitude and deep are correlated negatively with surf only for male gender.

Points correlate positively with attitude meaning that points and attitude increase together.
When gender is taken into account correlation is significant both for males and females.

## Regression
First, select three variables which correlate highest with the points to be used as explanatory variables. Those are: attitude, stra, and surf.

### Multiple regression analysis
Create a plot matrix and then a regression model having multiple explanatory variables.
```{r}
# create an plot matrix with ggpairs()
ggpairs(students2014, lower = list(combo = wrap("facethist", bins = 20)))

# create a regression model with multiple explanatory variables
multiple_variables <- lm(points ~ attitude + stra + surf, data = students2014)

# print out a summary of the model
summary(multiple_variables)
```
Inspecting the results show that points correlate in statistically significant manner only with variable attitude.
Let's remove the variables stra and surf from the model and use simple regression for variable attitude.
Will do simple regression for attitude.

### Simple regression
```{r}
# a scatter plot of points versus attitude
qplot(attitude, points, data = students2014) + geom_smooth(method = "lm")

# fit a linear model
linear_attitude <- lm(points ~ attitude, data = students2014)

# print out a summary of the model
summary(linear_attitude)
```
Inspecting the results show that adjusted R-squared is 0.19 indicating that attitude can explain about 19 % of the change in points. 


Now, to check the assumptions we'll produce diagnostics plots: Residuals vs Fitted values, Normal QQ-plot and Residuals vs Leverage.
```{r}
# draw diagnostic plots using the plot() function. Choose the plots 1, 2 and 5; Residuals vs Fitted values, Normal QQ-plot, and Residuals vs Leverage
plot(linear_attitude, which = c(1,2,5))
```

#### Interpretation of the diagnostic plots

**Residuals vs Fitted plot**

The line stays close to 0 indicating that the model works nicely for points up to 24 and less well for the points in the higher end. 
Observations 35, 56, and 145 are suggested as potential outliers.

**Normal Q-Q plot**

This probability plot indicates there is a linear connection and that the model fits nicely. For the low and high end the model fit is not. 
Observations 35, 56, and 145 are marked as possible outliers. 

**Residuals vs Leverage plot**

The line stays quite close to 0, which indicates there are no major outliers in the data. 
Observations 35, 56, and 71 are marked as potential outliers.


Linear model is expected to have a good fit on the data. These diagnostic plots support the choice of linear model.
